Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-02-25 paper/main.tex
progress in section 3 (and improved prelim again)

\newcommand{\Nset}{{\mathbb{N}_0}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
%\def\Inf{\ensuremath{\mathsf{Inf}}}
%\def\Fin{\ensuremath{\mathsf{Fin}}}
\DeclareMathOperator{\Inf}{\mathsf{Inf}}
\DeclareMathOperator{\Fin}{\mathsf{Fin}}
\newcommand{\minf}{\mathit{inf}}
\newcommand{\mks}{\mathit{mks}}
\def\false{\mathit{false}}
\def\true{\mathit{true}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\mB}{\mathcal{B}}

@@

A tuple $t=(p,a,M',q)\in\delta$ is the \emph{transition} leading from
state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$.  A
run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $\minf(\pi)$ denote the set of transitions that appear
infinitely many times in $\pi$.
%
We use $\mks(\cdot)$ to denote the set of acceptance
marks of various objects:
\begin{itemize}
\item for a transitions $t=(p,a,M',q)\in\delta$, $\mks(t)$ denotes the set $M'$,
\item for a set $T\subseteq\delta$ of transitions,
  $\mks(T)=\bigcup_{t\in T}\mks(t)$ denotes the set of marks that appear on
  transitions in $T$,
\item for a run $\pi$, $\mks(\pi)=\mks(\minf(\pi))$ denotes the set of
  marks that occur in infinitely many transitions of run $\pi$.
\end{itemize}
%
Run $\pi$ is \emph{accepting} iff $\mks(\pi)$ satisfies the formula
$\varphi$, where a set of marks $M'\subseteq M$ satisfies $\Inf m$ iff
$m\in M'$ and it satisfies $\Fin m$ iff $m\not\in M'$. The
\emph{language} of automaton $\mA$ is the set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

@@

A \emph{path} from a state $p$ to a state $q$ is a finite sequence of
adjacent transitions of the form
$(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. We assume that each state $q$ is
reachable from the initial state $q_I$ (i.e., $q=q_I$ or there is a
path from $q_I$ to $q$) as states that are not reachable from $q_I$
can be eliminated without any impact on $L(\mA)$. A nonempty set of
states $S\subseteq Q$ is called a nontrivial \emph{strongly connected
  component (SCC)} if for each $p,q\in S$ there is a path from $p$ to
$q$. An SCC S is \emph{maximal} if there is no SCC $S'$ satisfying
$S\subsetneq S'$. In the rest of this paper, SCC always refers to a
maximal SCC. Given an SCC $S$,
$\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$ denotes the
set of all transitions between states in $S$. Further, for each mark
$m\in M$, let $\delta_m$ denotes the set of all transitions marked
with $m$, i.e., $\delta_m=\{(p,a,M',q)\in\delta\mid m\in M'\}$. The
sets $\delta_m$ are called \emph{acceptance sets}.  Given an SCC $S$
and a mark $m\in M$, by $\delta_{S,m}$ we denote the set
$\delta_{S,m}=\delta_S\cap\delta_m$. Finally, a set of transitions
$C\subseteq\delta$ is called a \emph{cycle} is there exists a path
from a state $p$ to the same state containing all transitions of $C$
and no transition outside $C$.

@@

Recall that we aim to reduce the number of acceptance marks in a given
TELA $\mA =(Q,M,\Sigma,q_I,\delta,\varphi)$ without altering its
structure and language. In other words, we change only the acceptance
marks on individual transitions and we replace the acceptance formula
by a new one with less acceptance marks. To ensure that the modified
automaton is equivalent to the original one, we actually construct the
new acceptance formula such that the set of acceptance runs stays the
same (modulo changed acceptance marks on transitions).

We start with the several simple observations. Acceptance of a run
$\pi$ is fully determined by the set $M(\pi)$ of acceptance marks that
appear infinitely often on $\pi$, which is in turn fully determined by
the set $\minf(\pi)$ of transitions appearing infinitely often in
$\pi$.
% \[
%   M(\pi)=\bigcup_{(q_i,u_i,M_i,q_{i+1})\in\minf(\pi)}M_i
% \]
Further, each set of the form $\minf(\pi)$ is a cycle and vice versa.
Hence, our reduction methods is looking for a new acceptance formula
$\psi$ with less accceptance marks than the original acceptance
formula $\varphi$ and for new placement of the marks on transitions
such that for each cycle $C\subseteq\delta$ it holds that the marks $M(C)$
satisfies $\varphi$ if and only if $M(C)$ satisfies $\psi$.

>>>>

\newcommand{\Nset}{{\mathbb{N}_0}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
%\def\Inf{\ensuremath{\mathsf{Inf}}}
%\def\Fin{\ensuremath{\mathsf{Fin}}}
\DeclareMathOperator{\Inf}{\mathsf{Inf}}
\DeclareMathOperator{\Fin}{\mathsf{Fin}}
\newcommand{\minf}{\mathit{inf}}
\newcommand{\mks}{\mathit{mks}}
%\newcommand{\oM}{\widehat{M}}
\newcommand{\oM}{N}
\newcommand{\rem}{\mathit{nm}}
\newcommand{\mcycle}{\mathit{cycle}}
\newcommand{\msat}{\mathit{satisfies}}
\def\false{\mathit{false}}
\def\true{\mathit{true}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\mB}{\mathcal{B}}

@@

A tuple $t=(p,a,M',q)\in\delta$ is the \emph{transition} leading from
state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$.
The set $M'$ is also refered by $\mks(t)$. For a set of transitions
$T\subseteq\delta$, let $\mks(T)=\bigcup_{t\in T}\mks(t)$ denote the
set of marks that appear on transitions in $T$.
% of various objects:
% \begin{itemize}
% \item for a transition $t=(p,a,M',q)\in\delta$, $\mks(t)$ denotes the set $M'$,
% \item for a set $T\subseteq\delta$ of transitions,
%   $\mks(T)=\bigcup_{t\in T}\mks(t)$ denotes the set of marks that appear on
%   transitions in $T$,
% \item for a run $\pi$, $\mks(\pi)=\mks(\minf(\pi))$ denotes the set of
%   marks that occur in infinitely many transitions of run $\pi$.
% \end{itemize}

A run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $\minf(\pi)$ denote the set of transitions that appear
infinitely many times in $\pi$. Run $\pi$ is \emph{accepting} iff
$\minf(\pi)$ satisfies the formula $\varphi$, where a set of $T$ of
transitions satisfies $\Inf m$ iff $m\in\mks(T)$ and it satisfies
$\Fin m$ iff $m\not\in\mks(T)$. The \emph{language} of automaton $\mA$ is
the set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

@@

A \emph{path} from a state $p$ to a state $q$ is a finite sequence of
adjacent transitions of the form
$(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. A nonempty set of states $S\subseteq Q$
is called a nontrivial \emph{strongly connected component (SCC)} if
for each $p,q\in S$ there is a path from $p$ to $q$. An SCC S is
\emph{maximal} if there is no SCC $S'$ satisfying $S\subsetneq S'$. In
the rest of this paper, SCC always refers to a maximal SCC. Given an
SCC $S$, $\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$
denotes the set of all transitions between states in $S$. Further, for
each mark $m\in M$, let $\delta_m$ denotes the set of all transitions
marked with $m$, i.e., $\delta_m=\{t\in\delta\mid m\in
\mks(t)\}$. \todo{The sets $\delta_m$ are called \emph{acceptance
    sets}. Given an SCC $S$ and a mark $m\in M$, by $\delta_{S,m}$ we
  denote the set $\delta_{S,m}=\delta_S\cap\delta_m$.} A set of
transitions $T\subseteq\delta$ is called a \emph{cycle} is there
exists a path from a state $p$ to the same state containing all
transitions of $T$ and no transition outside $T$. Finally, we assume
that each TELA $\mA$ contains only states $q$ that are reachable from
the initial state $q_I$ (i.e., $q=q_I$ or there is a path from $q_I$
to $q$) as states that are not reachable from $q_I$ can be eliminated
without any impact on $L(\mA)$.

@@

Recall that we aim to reduce the number of acceptance marks in a given
TELA $\mA =(Q,M,\Sigma,q_I,\delta,\varphi)$ without altering its
structure and language. In other words, we are looking for a set $\oM$
of acceptance marks, an accepance formula $\psi$ over $\oM$, and a
function
$\rem:\delta\rightarrow Q \times \Sigma \times 2^{\oM}\times Q$
assigning new marks to transitions (i.e., for each
$t=(p,a,M',q)\in\delta$, we assume that $\rem(t)=(p,a,\oM',q)$ for
some $\oM'\subseteq\oM$) such that $|\oM|<|M|$ and the automaton
$\mB=(Q,\oM,\Sigma,q_I,\rem(\delta),\psi)$ is equivalent to $\mA$. To
ensure that the modified automaton is equivalent to the original one,
we actually look for $\psi$ and $\rem$ such that each run
$\pi=t_0t_1t_2\ldots$ of $\mA$ is accepting if and only if the run
$\rem(t_0)\rem(t_1)\rem(t_2)\ldots$ of $\mB$ is accepting.

We start with several simple observations. Acceptance of a run $\pi$
is fully determined by the set $\minf(\pi)$.
% of acceptance marks that
% appear infinitely often on $\pi$, which is in turn fully determined by
% the set $\minf(\pi)$ of transitions appearing infinitely often in
% $\pi$.
Further, each set $\minf(\pi)$ for a run $\pi$ is a cycle and
vice versa. Hence, our reduction methods is looking for a new
acceptance formula $\psi$ and function $\rem$ such that for each cycle
$T\subseteq\delta$, it holds that $T$ satisfies $\varphi$ if and
only if $\rem(T))$ satisifes $\psi$. This can be formally written
by the formula
\[
  \forall T\subseteq\delta~.~\mcycle(T)\implies(T~\msat~\varphi ~\iff~ \rem(T)~\msat~\psi).
\]
